---
title: "PBS Utah Donor Project: Exploratory Data Analysis"
author: "Owen Simon"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
execute:
  message: false
  warning: false
---

## Codex Session

codex resume 01a10e2e-1a08-7650-949e-773c161f08e8

## Purpose and approach

This report provides reproducible exploratory data analysis for the synthetic PBS Utah donor release. It documents file grain, data quality, distributions, time coverage, and relationships before feature engineering or modeling. All findings describe synthetic data only.

The source files are large. Each section loads one file, saves only small summaries, and removes the raw object before the next section. This makes rendering practical without keeping all nine files in memory.

## Setup

```{r}
# Load only packages required to import, summarize, and plot the project data.
library(dplyr)
library(gt)
library(ggplot2)
library(knitr)
library(readr)
library(tibble)

# Use one relative path so the report can run from the project root.
data_dir <- "Data_Files"

# Read a data file quietly and consistently.
read_project_csv <- function(file_name) {
  read_csv(file.path(data_dir, file_name), show_col_types = FALSE, progress = FALSE)
}

# Calculate PBS fiscal year, where July--December belong to the next fiscal year.
fiscal_year <- function(date) {
  as.integer(format(date, "%Y")) + as.integer(format(date, "%m") >= 7)
}

# Summarize a table without printing or retaining its raw records.
profile_data <- function(data, file_name, key = NULL) {
  key_missing <- key_duplicates <- NA_integer_
  if (!is.null(key) && key %in% names(data)) {
    values <- data[[key]]
    key_missing <- sum(is.na(values) | values == "")
    key_duplicates <- sum(duplicated(values[!is.na(values) & values != ""]))
  }
  tibble(file = file_name, rows = nrow(data), columns = ncol(data),
         key = key,
         missing_key_values = key_missing, duplicate_nonmissing_keys = key_duplicates)
}

# Report blank strings and missing values for each field.
missingness <- function(data) {
  tibble(field = names(data),
         missing_n = vapply(data, function(x) sum(is.na(x) | x == ""), integer(1))) %>%
    mutate(missing_pct = round(100 * missing_n / nrow(data), 2)) %>%
    arrange(desc(missing_pct), desc(missing_n))
}

# Render compact summary tables with readable numbers.
show_table <- function(x) kable(x, format = "html", digits = 2,
                                format.args = list(big.mark = ","))
```

## Data inventory

The table grain is essential context for joins. A constituent, payment, legacy household gift, solicitation, soft credit, cultivation assignment, and viewing record are not interchangeable units.

```{r}
# Document all expected inputs and verify that they are present in the repository.
inventory <- tribble(
  ~file, ~grain, ~documented_key,
  "constituents_w_memberships.csv", "constituent", "Constituent ID",
  "unite_payments.csv", "payment", "payment_id",
  "team_approach_legacy_payments.csv", "legacy household gift", "ta_account_id",
  "soft_credits.csv", "soft credit", "Credit ID",
  "campaign_codes.csv", "marketing code reference", "Marketing Code",
  "campaign_members.csv", "solicitation", "none documented",
  "cultivation.csv", "constituent-year assignment", "synthetic_id plus fiscal_year",
  "officer_portfolios.csv", "portfolio", "portfolio_id",
  "passport_viewing.csv", "viewing record", "UID"
) %>%
  mutate(available = file.exists(file.path(data_dir, file)),
         bytes = if_else(available, file.info(file.path(data_dir, file))$size, NA_real_))
```

```{r}
gt(inventory) |>
  tab_header(
    title = "Data File Library")
```

## Constituents and memberships

This entity table supplies donor attributes. First check the documented identifier, missingness, demographic distributions, and current membership attributes. Current snapshot variables must not be used as historical values.

```{r}
# Load the entity table and create only summaries needed for the report.
constituents <- read_project_csv("constituents_w_memberships.csv")

constituent_profile <- profile_data(constituents, "constituents_w_memberships.csv", "Constituent ID")

constituent_missing <- missingness(constituents)

constituent_summary <- constituents %>%
  summarise(unique_ids = n_distinct(.data[["Constituent ID"]]),
            median_age = median(Age, na.rm = TRUE),
            sustainers = sum(.data[["Sustainer PBS Utah"]], na.rm = TRUE),
            sustainer_pct = 100 * mean(.data[["Sustainer PBS Utah"]], na.rm = TRUE))

segments <- constituents %>%
  count(.data[["Primary Constituent Type"]], Gender, .data[["Sustainer PBS Utah"]], sort = TRUE)
```

```{r}
gt(constituent_profile)
gt(constituent_summary) |>
  tab_header(
    title = ""
  )

gt(head(constituent_missing, 12))
gt(head(segments, 15))
```

```{r}
# Plot the age distribution only for records with known ages.
ggplot(filter(constituents, !is.na(Age)), aes(Age)) +
  geom_histogram(binwidth = 5, fill = "#2563A6", color = "white") +
  labs(title = "Constituent age distribution", x = "Age", y = "Constituents") +
  theme_minimal()
```

```{R}
# Release large raw data before analyzing transactions.
rm(constituents, constituent_profile, constituent_missing, constituent_summary, segments)
gc()
```

## Unite payments

Unite contains payment-level data beginning in FY2021. Examine key quality, invalid values, donor coverage, fiscal-year trend, and mix of gift and tender types. Revenue totals and transaction counts answer different questions and are both shown.

```{r}
# Parse the date explicitly before grouping payments into fiscal years.
unite <- read_project_csv("unite_payments.csv") %>%
  mutate(credit_date = as.Date(credit_date), 
         fiscal_year = fiscal_year(credit_date))

unite_profile <- profile_data(unite, "unite_payments.csv", "payment_id")

unite_quality <- unite %>%
  summarise(unique_donors = n_distinct(donor_id),
            nonpositive_amounts = sum(is.na(payment_amount) | payment_amount <= 0),
            missing_dates = sum(is.na(credit_date)),
            first_date = min(credit_date, na.rm = TRUE), last_date = max(credit_date, na.rm = TRUE))

unite_by_year <- unite %>%
  group_by(fiscal_year) %>%
  summarise(payments = n(), 
            donors = n_distinct(donor_id), 
            revenue = sum(payment_amount, na.rm = TRUE),
            median_payment = median(payment_amount, na.rm = TRUE), .groups = "drop")

unite_mix <- unite %>%
  group_by(gift_type, tender_type) %>%
  summarise(payments = n(), revenue = sum(payment_amount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(revenue))
```

```{r}
gt(unite_profile)
gt(unite_quality) |>
  tab_header(
    title = "Unite Data Quality")

gt(unite_by_year)
gt(head(unite_mix, 15))
```

```{r}
# Plot annual revenue and investigate discontinuities before drawing conclusions.
ggplot(unite_by_year, aes(factor(fiscal_year), revenue)) +
  geom_col(fill = "#2563A6") +
  scale_y_continuous(labels = function(x) paste0("$", format(x, big.mark = ",", trim = TRUE))) +
  labs(title = "Unite payment revenue by fiscal year", x = "Fiscal year", y = "Revenue") +
  theme_minimal()

rm(unite, unite_profile, unite_quality, unite_by_year, unite_mix)
gc()
```

## Legacy Team Approach payments

Legacy records have household-gift grain and can link to several constituent IDs. Do not append their rows to Unite and call the result donor-level data. First inspect link availability, time range, values, and communication channels; later aggregate to an explicit donor-year rule.

```{r}
# Parse legacy dates and quantify constituent-link completeness.
legacy <- read_project_csv("team_approach_legacy_payments.csv") %>%
  mutate(gift_date = as.Date(gift_date), fiscal_year = fiscal_year(gift_date))

legacy_quality <- legacy %>%
  summarise(nonpositive_amounts = sum(is.na(payment_amount) | payment_amount <= 0),
            missing_dates = sum(is.na(gift_date)),
            first_constituent_link = sum(!is.na(constituent_id_1) & constituent_id_1 != ""),
            no_constituent_link = sum(if_all(starts_with("constituent_id_"), ~ is.na(.x) | .x == "")))

legacy_year <- legacy %>%
  group_by(fiscal_year) %>%
  summarise(gifts = n(), revenue = sum(payment_amount, na.rm = TRUE),
            median_gift = median(payment_amount, na.rm = TRUE), .groups = "drop")

legacy_channels <- legacy %>%
  group_by(communication_method, communication_type) %>%
  summarise(gifts = n(), revenue = sum(payment_amount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(revenue))
```

```{r}
gt(profile_data(legacy, "team_approach_legacy_payments.csv", "ta_account_id"))
gt(legacy_quality)
gt(tail(legacy_year, 12))
gt(head(legacy_channels, 15))
```

```{r}
rm(legacy, legacy_quality, legacy_year, legacy_channels)
gc()
```

## Campaign codes and solicitation history

Campaign codes are reference data; campaign members are solicitation records. The anti-join checks whether every solicitation points to a known code. Response rates are descriptive, not causal measures of campaign effectiveness.

```{r}
# Load both campaign tables only while their relationship is being tested.
campaign_codes <- read_project_csv("campaign_codes.csv")

campaign_members <- read_project_csv("campaign_members.csv") %>%
  mutate(campaign_start_date = as.Date(campaign_start_date),
         fiscal_year = fiscal_year(campaign_start_date))

campaign_links <- campaign_members %>%
  anti_join(select(campaign_codes, marketing_code = all_of("Marketing Code")), by = "marketing_code") %>%
  summarise(unmatched_rows = n(), unmatched_codes = n_distinct(marketing_code))

campaign_results <- campaign_members %>%
  group_by(fiscal_year) %>%
  summarise(solicitations = n(), constituents = n_distinct(constituent_id),
            responses = sum(responded, na.rm = TRUE),
            response_rate_pct = 100 * mean(responded, na.rm = TRUE), .groups = "drop")

campaign_rank <- campaign_members %>%
  group_by(marketing_code, campaign_name) %>%
  summarise(solicitations = n(), responses = sum(responded, na.rm = TRUE),
            response_rate_pct = 100 * mean(responded, na.rm = TRUE), .groups = "drop") %>%
  filter(solicitations >= 30) %>% arrange(desc(response_rate_pct), desc(solicitations))
```

```{r}
gt(profile_data(campaign_codes, "campaign_codes.csv", "Marketing Code"))
gt(profile_data(campaign_members, "campaign_members.csv"))
gt(campaign_links)
gt(campaign_results)
gt(head(campaign_rank, 20))
```

```{r}
rm(campaign_codes, campaign_members, campaign_links, campaign_results, campaign_rank)
gc()
```

## Soft credits

Soft credits are attributed recognition, rather than necessarily direct cash paid by the recipient. They should stay separate from payments until a later metric applies the documented duplicate-removal and attribution rules.

```{r}
# Examine unique credits, orphan recipients, amount validity, and organization type.
soft <- read_project_csv("soft_credits.csv") %>%
  mutate(credit_date = as.Date(.data[["Credit Date"]]), fiscal_year = fiscal_year(credit_date))
soft_quality <- soft %>%
  summarise(exact_duplicate_rows = sum(duplicated(soft)),
            orphan_recipients = sum(.data[["Soft Credit Constituent ID"]] == "SYN-ORPHAN", na.rm = TRUE),
            nonpositive_amounts = sum(is.na(.data[["Credit Amount"]]) | .data[["Credit Amount"]] <= 0),
            first_date = min(credit_date, na.rm = TRUE), last_date = max(credit_date, na.rm = TRUE))
soft_mix <- soft %>%
  group_by(.data[["Hard Credit Org Type"]], .data[["Credit Gift Type"]]) %>%
  summarise(credits = n(), amount = sum(.data[["Credit Amount"]], na.rm = TRUE),
            median_amount = median(.data[["Credit Amount"]], na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(amount))
```

```{r}
gt(profile_data(soft, "soft_credits.csv", "Credit ID"))
gt(soft_quality)
gt(soft_mix)
```

```{r}
rm(soft, soft_quality, soft_mix)
gc()
```

## Cultivation assignments and officer portfolios

Cultivation has constituent-year grain while portfolios are reference records. The following relationship check and funnel help assess capacity, selection, contact, and holdout patterns. Holdout comparisons require further validation before causal interpretation.

```{r}
# Join portfolio capacity to cultivation summaries and report unmatched assignments.
portfolios <- read_project_csv("officer_portfolios.csv")

cultivation <- read_project_csv("cultivation.csv")

cultivation_links <- cultivation %>%
  anti_join(portfolios, by = "portfolio_id") %>%
  summarise(unmatched_rows = n(), unmatched_portfolios = n_distinct(portfolio_id))

cultivation_funnel <- cultivation %>%
  group_by(fiscal_year, portfolio_id) %>%
  summarise(assignments = n(), selected = sum(selected, na.rm = TRUE),
            contacted = sum(contacted, na.rm = TRUE), holdout = sum(holdout, na.rm = TRUE),
            adopted = first(program_adopted), .groups = "drop") %>%
  left_join(select(portfolios, portfolio_id, capacity, rollout_fy), by = "portfolio_id") %>%
  mutate(selection_rate_pct = 100 * selected / assignments,
         contact_rate_of_selected_pct = if_else(selected > 0, 100 * contacted / selected, NA_real_))
```

```{r}
gt(profile_data(portfolios, "officer_portfolios.csv", "portfolio_id"))
gt(profile_data(cultivation, "cultivation.csv"))
gt(cultivation_links)
gt(cultivation_funnel)
```

```{r}
rm(portfolios, cultivation, cultivation_links, cultivation_funnel)
gc()
```

## Passport viewing data

The checked-in Passport file may be a Git LFS pointer rather than data. This chunk reports that condition cleanly. After retrieving the LFS object, it will profile account reach, viewing time, completion, content, and device distribution.

```{r}
# Import Passport viewing data
passport <- read_project_csv("passport_viewing.txt")

passport_summary <- passport %>%
  summarise(
    records = n(),
    linked_constituents = n_distinct(constituent_id, na.rm = TRUE),
    total_titles_viewed = sum(titles_viewed, na.rm = TRUE),
    median_completion_pct = median(mean_percent_watched, na.rm = TRUE)
  )

passport_content <- passport %>%
  group_by(fiscal_year, genre) %>%
  summarise(
    constituents = n_distinct(constituent_id, na.rm = TRUE),
    titles_viewed = sum(titles_viewed, na.rm = TRUE),
    mean_completion_pct = mean(mean_percent_watched, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(constituents), desc(titles_viewed))
```

```{r}
gt(profile_data(passport, "passport_viewing.txt", "constituent_id"))
gt(passport_summary)
gt(head(passport_content, 20))
```

```{r}
rm(passport, passport_summary, passport_content)
gc()
```

## EDA conclusions and next steps

Read the resulting tables before building features. Resolve unexpected duplicate keys, broken links, non-positive amounts, and missing dates. Define a single fiscal-year giving outcome using the documented direct-payment and soft-credit rules. Aggregate every source to an explicit analysis grain before joining it. Treat response rates and cultivation comparisons as descriptive until eligibility, selection, timing, and holdout design are confirmed. Passport nonappearance does not prove no viewing or account access.

These checks are intended to become automated validation tests in the feature-engineering workflow.
