---
title: "PBS Utah Donor Project: Exploratory Data Analysis"
author: "Owen Simon"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
execute:
  message: false
  warning: false
---

## Setup

```{r}
# Load only packages required to import, summarize, and plot the project data.
library(dplyr)
library(gt)
library(ggplot2)
library(knitr)
library(readr)
library(tibble)

# Use one relative path so the report can run from the project root.
data_dir <- "Data_Files"

# Read a data file quietly and consistently.
read_project_csv <- function(file_name) {
  read_csv(file.path(data_dir, file_name), show_col_types = FALSE, progress = FALSE)
}

# Calculate PBS fiscal year, where July--December belong to the next fiscal year.
fiscal_year <- function(date) {
  as.integer(format(date, "%Y")) + as.integer(format(date, "%m") >= 7)
}

# Summarize a table without printing or retaining its raw records.
profile_data <- function(data, file_name, key = NULL) {
  key_missing <- key_duplicates <- NA_integer_
  if (!is.null(key) && key %in% names(data)) {
    values <- data[[key]]
    key_missing <- sum(is.na(values) | values == "")
    key_duplicates <- sum(duplicated(values[!is.na(values) & values != ""]))
  }
  tibble(file = file_name, rows = nrow(data), columns = ncol(data),
         key = key,
         missing_key_values = key_missing, duplicate_nonmissing_keys = key_duplicates)
}

# Report blank strings and missing values for each field.
missingness <- function(data) {
  tibble(field = names(data),
         missing_n = vapply(data, function(x) sum(is.na(x) | x == ""), integer(1))) %>%
    mutate(missing_pct = round(100 * missing_n / nrow(data), 2)) %>%
    arrange(desc(missing_pct), desc(missing_n))
}

# Render compact summary tables with readable numbers.
show_table <- function(x) kable(x, format = "html", digits = 2,
                                format.args = list(big.mark = ","))
```

## Data inventory

The table grain is essential context for joins. A constituent, payment, legacy household gift, solicitation, soft credit, cultivation assignment, and viewing record are not interchangeable units.

```{r}
# Document all expected inputs and verify that they are present in the repository.
inventory <- tribble(
  ~file, ~grain, ~documented_key,
  "constituents_w_memberships.csv", "constituent", "Constituent ID",
  "unite_payments.csv", "payment", "payment_id",
  "team_approach_legacy_payments.csv", "legacy household gift", "ta_account_id",
  "soft_credits.csv", "soft credit", "Credit ID",
  "campaign_codes.csv", "marketing code reference", "Marketing Code",
  "campaign_members.csv", "solicitation", "none documented",
  "cultivation.csv", "constituent-year assignment", "synthetic_id plus fiscal_year",
  "officer_portfolios.csv", "portfolio", "portfolio_id",
  "passport_viewing.csv", "viewing record", "UID"
) %>%
  mutate(available = file.exists(file.path(data_dir, file)),
         bytes = if_else(available, file.info(file.path(data_dir, file))$size, NA_real_))
```

```{r}
gt(inventory) |>
  tab_header(
    title = "Data File Library")
```

## Constituents and memberships

This entity table supplies donor attributes. First check the documented identifier, missingness, demographic distributions, and current membership attributes. Current snapshot variables must not be used as historical values.

```{r}
# Load the entity table and create only summaries needed for the report.
constituents <- read_project_csv("constituents_w_memberships.csv")

constituent_profile <- profile_data(constituents, "constituents_w_memberships.csv", "Constituent ID")

constituent_missing <- missingness(constituents)

constituent_summary <- constituents %>%
  summarise(unique_ids = n_distinct(.data[["Constituent ID"]]),
            median_age = median(Age, na.rm = TRUE),
            sustainers = sum(.data[["Sustainer PBS Utah"]], na.rm = TRUE),
            sustainer_pct = 100 * mean(.data[["Sustainer PBS Utah"]], na.rm = TRUE))

segments <- constituents %>%
  count(.data[["Primary Constituent Type"]], Gender, .data[["Sustainer PBS Utah"]], sort = TRUE)
```

```{r}
gt(constituent_profile) |>
  tab_header(
    title = "Constituent Data Profile"
  )
```

196,167 rows; no duplicates; no missing key values

```{r}
gt(constituent_summary) |>
  tab_header(
    title = "Constituent Summary"
  )
```

```{r}
# Plot the age distribution only for records with known ages.
ggplot(filter(constituents, !is.na(Age)), aes(Age)) +
  geom_histogram(binwidth = 5, fill = "#2563A6", color = "white") +
  labs(title = "Constituent age distribution", x = "Age", y = "Constituents") +
  theme_minimal()
```

The median constituent age is 68 years old; 13.92% of constituents are PBS Utah sustainers (giving a monthly recurring gift).

```{r}
gt(head(constituent_missing, 12))
```

Variables with nearly 100% missing values:

- UofU Degree Info
- Latest UofU Degree Year
- UofU Employment Status
- Major Donor Class PBS Utah
- Spouse Constituent ID
- Marital Status

```{r}
gt(head(segments, 15))
```

```{r}
#| echo: False
cat("'Friends' Primary Constituent Type makes up", round(100 * (sum(segments$n[segments$`Primary Constituent Type` == "Friends"]) / sum(segments$n)), 2), "percent of the data.")
```

```{r}
#| include: False
# Release large raw data before analyzing transactions.
rm(constituents, constituent_profile, constituent_missing, constituent_summary, segments)
gc()
```

## Unite payments

Unite contains payment-level data beginning in FY2021. Examine key quality, invalid values, donor coverage, fiscal-year trend, and mix of gift and tender types. Revenue totals and transaction counts answer different questions and are both shown.

```{r}
# Parse the date explicitly before grouping payments into fiscal years.
unite <- read_project_csv("unite_payments.csv") %>%
  mutate(credit_date = as.Date(credit_date), 
         fiscal_year = fiscal_year(credit_date))

unite_profile <- profile_data(unite, "unite_payments.csv", "payment_id")

unite_quality <- unite %>%
  summarise(unique_donors = n_distinct(donor_id),
            nonpositive_amounts = sum(is.na(payment_amount) | payment_amount <= 0),
            missing_dates = sum(is.na(credit_date)),
            first_date = min(credit_date, na.rm = TRUE), last_date = max(credit_date, na.rm = TRUE))

unite_by_year <- unite %>%
  group_by(fiscal_year) %>%
  summarise(payments = n(), 
            donors = n_distinct(donor_id), 
            revenue = sum(payment_amount, na.rm = TRUE),
            median_payment = median(payment_amount, na.rm = TRUE), .groups = "drop")

unite_mix <- unite %>%
  group_by(gift_type, tender_type) %>%
  summarise(payments = n(), revenue = sum(payment_amount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(revenue))
```

```{r}
gt(unite_profile)|>
  tab_header(
    title = "Unite Data Profile"
  )
```

1,722,698 rows; no duplicates; no missing key values

```{r}
gt(unite_quality) |>
  tab_header(
    title = "Unite Data Quality")
```

71,643 unique donors; 1,086 non-positive unite donations; no missing dates bewteen July 01, 2020 and June 30, 2026.

```{r}
gt(unite_by_year)
```

```{r}
# Plot annual revenue and investigate discontinuities before drawing conclusions.
ggplot(unite_by_year, aes(factor(fiscal_year), revenue)) +
  geom_col(fill = "#2563A6") +
  scale_y_continuous(labels = function(x) paste0("$", format(x, big.mark = ",", trim = TRUE))) +
  labs(title = "Unite payment revenue by fiscal year", x = "Fiscal year", y = "Revenue") +
  theme_minimal()
```

```{r}
#| echo: False
cat("$", format(sum(unite_by_year$revenue, na.rm = TRUE), big.mark = ",", scientific = FALSE), " of revenue generated since 2020.\n", sep = "")

highest <- unite_by_year[which.max(unite_by_year$revenue), ]
lowest <- unite_by_year[which.min(unite_by_year$revenue), ]

cat("Highest revenue: $", format(highest$revenue, big.mark = ","),
    " in fiscal year ", highest$fiscal_year, ".\n", sep = "")

cat("Lowest revenue: $", format(lowest$revenue, big.mark = ","),
    " in fiscal year ", lowest$fiscal_year, ".\n", sep = "")
```

```{r}
gt(head(unite_mix, 10))
```

```{r}
#| echo: False
tender_summary <- unite_mix |>
  group_by(tender_type) |>
  summarise(revenue = sum(revenue, na.rm = TRUE))

highest <- tender_summary |> slice_max(revenue, n = 1)
lowest <- tender_summary |> slice_min(revenue, n = 1)

cat("Highest revenue: $", format(highest$revenue, big.mark = ","),
    " from tender type ", highest$tender_type, ".\n", sep = "")

cat("Lowest revenue: $", format(lowest$revenue, big.mark = ","),
    " from tender type ", lowest$tender_type, ".\n", sep = "")
```

```{r}
#| include: False
rm(unite, unite_profile, unite_quality, unite_by_year, unite_mix)
gc()
```

## Legacy Team Approach payments

Legacy records have household-gift grain and can link to several constituent IDs. Do not append their rows to Unite and call the result donor-level data. First inspect link availability, time range, values, and communication channels; later aggregate to an explicit donor-year rule.

```{r}
# Parse legacy dates and quantify constituent-link completeness.
legacy <- read_project_csv("team_approach_legacy_payments.csv") %>%
  mutate(gift_date = as.Date(gift_date), fiscal_year = fiscal_year(gift_date))

legacy_quality <- legacy %>%
  summarise(nonpositive_amounts = sum(is.na(payment_amount) | payment_amount <= 0),
            missing_dates = sum(is.na(gift_date)),
            first_constituent_link = sum(!is.na(constituent_id_1) & constituent_id_1 != ""),
            no_constituent_link = sum(if_all(starts_with("constituent_id_"), ~ is.na(.x) | .x == "")))

legacy_year <- legacy %>%
  group_by(fiscal_year) %>%
  summarise(gifts = n(), revenue = sum(payment_amount, na.rm = TRUE),
            median_gift = median(payment_amount, na.rm = TRUE), .groups = "drop")

legacy_channels <- legacy %>%
  group_by(communication_method, communication_type) %>%
  summarise(gifts = n(), revenue = sum(payment_amount, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(revenue))
```

```{r}
gt(profile_data(legacy, "team_approach_legacy_payments.csv", "ta_account_id"))
```

```{r}
#| echo: False
cat("Unique rows:", length(unique(legacy$ta_account_id)))
```

```{r}
gt(legacy_quality)
```

```{r}
gt(tail(legacy_year, 12))
```

```{r}
gt(head(legacy_channels, 15))
```

```{r}
#| include: False
rm(legacy, legacy_quality, legacy_year, legacy_channels)
gc()
```

## Campaign Codes and Solicitation History

Campaign codes are reference data; campaign members are solicitation records. The anti-join checks whether every solicitation points to a known code. Response rates are descriptive, not causal measures of campaign effectiveness.

```{r}
# Load both campaign tables only while their relationship is being tested.
campaign_codes <- read_project_csv("campaign_codes.csv")

campaign_members <- read_project_csv("campaign_members.csv") %>%
  mutate(campaign_start_date = as.Date(campaign_start_date),
         fiscal_year = fiscal_year(campaign_start_date))

campaign_results <- campaign_members %>%
  group_by(fiscal_year) %>%
  summarise(solicitations = n(), constituents = n_distinct(constituent_id),
            responses = sum(responded, na.rm = TRUE),
            response_rate_pct = 100 * mean(responded, na.rm = TRUE), .groups = "drop")

campaign_rank <- campaign_members %>%
  group_by(marketing_code, campaign_name) %>%
  summarise(solicitations = n(), responses = sum(responded, na.rm = TRUE),
            response_rate_pct = 100 * mean(responded, na.rm = TRUE), .groups = "drop") %>%
  filter(solicitations >= 30) %>% arrange(desc(response_rate_pct), desc(solicitations))
```

```{r}
gt(profile_data(campaign_codes, "campaign_codes.csv", "Marketing Code"))|>
  tab_header(
    title = "Campaign Code Data Profile"
  )
```

29,249 rows; 0 missing; 0 duplicates

```{r}
gt(campaign_results)
```

```{r}
#| echo: False
cat("Average solicitation response rate:", round(mean(campaign_results$response_rate_pct), 2), "percent")
```

```{r}
gt(head(campaign_rank, 5))
```

Most effective campaign: Synthetic Campaign 000456

```{r}
gt(tail(campaign_rank, 5))
```

Least effective campaign: Synthetic Campaign 028114

```{r}
#| include: False
rm(campaign_codes, campaign_members, campaign_links, campaign_results, campaign_rank)
gc()
```

## Soft credits

Soft credits are attributed recognition, rather than necessarily direct cash paid by the recipient. They should stay separate from payments until a later metric applies the documented duplicate-removal and attribution rules.

```{r}
# Examine unique credits, orphan recipients, amount validity, and organization type.
soft <- read_project_csv("soft_credits.csv") %>%
  mutate(credit_date = as.Date(.data[["Credit Date"]]), fiscal_year = fiscal_year(credit_date))
soft_quality <- soft %>%
  summarise(exact_duplicate_rows = sum(duplicated(soft)),
            orphan_recipients = sum(.data[["Soft Credit Constituent ID"]] == "SYN-ORPHAN", na.rm = TRUE),
            nonpositive_amounts = sum(is.na(.data[["Credit Amount"]]) | .data[["Credit Amount"]] <= 0),
            first_date = min(credit_date, na.rm = TRUE), last_date = max(credit_date, na.rm = TRUE))
soft_mix <- soft %>%
  group_by(.data[["Hard Credit Org Type"]], .data[["Credit Gift Type"]]) %>%
  summarise(credits = n(), amount = sum(.data[["Credit Amount"]], na.rm = TRUE),
            median_amount = median(.data[["Credit Amount"]], na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(amount))
```

```{r}
gt(profile_data(soft, "soft_credits.csv", "Credit ID"))
```

```{r}
gt(soft_quality)
```

```{r}
gt(soft_mix)
```

```{r}
#| include: False
rm(soft, soft_quality, soft_mix)
gc()
```

## Cultivation assignments and officer portfolios

Cultivation has constituent-year grain while portfolios are reference records. The following relationship check and funnel help assess capacity, selection, contact, and holdout patterns. Holdout comparisons require further validation before causal interpretation.

```{r}
# Join portfolio capacity to cultivation summaries and report unmatched assignments.
portfolios <- read_project_csv("officer_portfolios.csv")

cultivation <- read_project_csv("cultivation.csv")

cultivation_links <- cultivation %>%
  anti_join(portfolios, by = "portfolio_id") %>%
  summarise(unmatched_rows = n(), unmatched_portfolios = n_distinct(portfolio_id))

cultivation_funnel <- cultivation %>%
  group_by(fiscal_year, portfolio_id) %>%
  summarise(assignments = n(), selected = sum(selected, na.rm = TRUE),
            contacted = sum(contacted, na.rm = TRUE), holdout = sum(holdout, na.rm = TRUE),
            adopted = first(program_adopted), .groups = "drop") %>%
  left_join(select(portfolios, portfolio_id, capacity, rollout_fy), by = "portfolio_id") %>%
  mutate(selection_rate_pct = 100 * selected / assignments,
         contact_rate_of_selected_pct = if_else(selected > 0, 100 * contacted / selected, NA_real_))
```

```{r}
gt(profile_data(portfolios, "officer_portfolios.csv", "portfolio_id"))
```

```{r}
cultivation_funnel |>
  filter(fiscal_year == 2025) |>
  gt() |>
    tab_header(
    title = "Latest Year Portfolios"
  )
```

```{r}
rm(portfolios, cultivation, cultivation_links, cultivation_funnel)
gc()
```

## Passport viewing data

The checked-in Passport file may be a Git LFS pointer rather than data. This chunk reports that condition cleanly. After retrieving the LFS object, it will profile account reach, viewing time, completion, content, and device distribution.

```{r}
# Import Passport viewing data
passport <- read_project_csv("passport_viewing.txt")

passport_summary <- passport %>%
  summarise(
    records = n(),
    linked_constituents = n_distinct(constituent_id, na.rm = TRUE),
    total_titles_viewed = sum(titles_viewed, na.rm = TRUE),
    median_completion_pct = median(mean_percent_watched, na.rm = TRUE)
  )

passport_content <- passport %>%
  group_by(fiscal_year, genre) %>%
  summarise(
    constituents = n_distinct(constituent_id, na.rm = TRUE),
    titles_viewed = sum(titles_viewed, na.rm = TRUE),
    mean_completion_pct = mean(mean_percent_watched, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(constituents), desc(titles_viewed))
```

```{r}
gt(profile_data(passport, "passport_viewing.txt", "constituent_id")) |>
    tab_header(
    title = "Passport Data Profile"
  )
```

```{r}
gt(passport_summary)
```

```{r}
gt(head(passport_content, 20))
```

```{r}
#| echo: False
genre_summary <- passport_content |>
  group_by(genre) |>
  summarise(titles_viewed = sum(titles_viewed, na.rm = TRUE))

highest <- genre_summary |> slice_max(titles_viewed, n = 1)
lowest <- genre_summary |> slice_min(titles_viewed, n = 1)

cat("Most titles viewed: ", format(highest$titles_viewed, big.mark = ","),
    " in genre ", highest$genre, ".\n", sep = "")

cat("Fewest titles viewed: ", format(lowest$titles_viewed, big.mark = ","),
    " in genre ", lowest$genre, ".\n", sep = "")
```

```{r}
#| include: False
rm(passport, passport_summary, passport_content)
gc()
```

